Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-11-28 paper/main.tex
rewrite text for benchmarking extraction algorithm

In order to determine if our extraction algorithm provides reliable results, we set out to compare
extracted data with known real data.
Our methodology makes use of the fact that we have already set up our model and solving scheme and
are able to produce cell masks from these numerical results.
In doing so, we have exact knowledge about the position of our agents since we can simply obtain
this information from our computational model.
But on the other hand we can also try to extract these positions from the generated cell-masks.
By comparing the differences between them, we get a good estimate of how well our extraction
algorithm can determine the true position of the bacteria.

In figure~\ref{fig:benchmarking-extraction-algorithm} (A-C) we see snapshots of a time series of
masks.
This particular simulation includes division events between images (B) and (C).
For all images (A-C), we have drawn the exact position of the agents with white lines and indicated
the extracted positions with crosses.
They align very well visually.
It may be hard to spot with the bare eye, but the agreement between extracted and exact positions is
visually worse at the endpoints of the rods.
To quantify this disparity, we calculate the distance between these individual vertices.
Figure~\ref{fig:benchmarking-extraction-algorithm} (D) shows the estimated length of the rods from
the extracted and exact vertices which is surrounded by the average vertex distance.
Division events lead to steps in our rod-length function since the bacterial length is halved.
We can clearly see that the agreement is very good in the beginning but right after every division
event, the uncertainty becomes large.
To illustrate this, we can see that in figure~\ref{fig:benchmarking-extraction-algorithm} (C) some
of the rods are partially overlapping which hides parts of the individual cell-masks and thus
changes the extraction result.
We also plotted the distribution of individual vertex distances in
figure~\ref{fig:benchmarking-extraction-algorithm} (E).
Darker areas in the plot correspond to distances at time-points earlier in the simulatin time.
We can see that brighter areas tend to shift towards the right, thus increasing the uncertainty of
the extraction.
However, the overall distribution shows small values of distances for a large number of vertices and
is not dominated by outliers.

\textbf{
    Die Verständlichkeit und die Formulierungen hier bedürfen noch der Verbesserung.
    Damit ich aber schneller durchkomme, verschieben wir das auf später.
}

>>>>

To validate the reliability of the extraction algorithm
(Algorithm~\ref{algo:extraction-algorithm}), we benchmarked it using synthetic data with known
ground truth.
This reference data in the form of vertices can be obtained directly from our computational model.
Furthermore as we have explained in section~\ref{subsection:visualization}, we generate cell masks
from a given set of vertices and thickness of the rod which will serve as inputs for our extraction
algorithm.
By comparing the extracted positions with known ground truth, we will obtain quantitative measures
of the extraction accuracy.

Figure~\ref{fig:benchmarking-extraction-algorithm} (D-H) shows snapshots of the synthetically
generated cell masks.
The initial configuration consists of 4 bacteria which grow and ultimately divide thus expanding
the filled up area.
Meanwhile figure~\ref{fig:benchmarking-extraction-algorithm} (I) shows a zoomed-in version of
snapshot (G), highlighting the direct comparison of the exact (dark cross) and extracted vertices.
Visual comparison of this time series indicates strong alignment between exact and extracted
results.
Upon closer inspection with figure~\ref{fig:benchmarking-extraction-algorithm} (I), we can see
that small differences start to manifest, especially at the tip of the rods in more crowded regions.

In order to quantitatively analyze this disparity we use both data to calculate the average rod
length of the bacteria (Figure~\ref{fig:benchmarking-extraction-algorithm} (J)) over the course of
the simulated time interval.
The results show good agreement with larger discrepancies directly after division events.
In addition figure~\ref{fig:benchmarking-extraction-algorithm} (K) displays individual displacements
of exact and extracted vertices which have been reoriented with respect to the direction of the rod.
Values closer to the center represent better agreement.
The distribution of these displacements can be well described by a normal distribution with mean
$\mu$ and variance $\sigma$.
These parameters can be obtained by fitting the distribution to the displacements at every time
point which is displayed in Figure~\ref{fig:benchmarking-extraction-algorithm}.
Here we can also see that the overall variance in the displacements increases directly after
division events.
However, the center of the distributions remain concentrated near zero without significant outliers
which demonstrates the robustness of our method in these densely packed scenarios.
Further details are given in the supplementary
materialsec:supsec:supplement-imaging-based-comparison-algorithm-details
(section~\ref{sec:supplement-imaging-based-comparison-algorithm-details}).
